---
title: 'Chapter 5: Linear Regression'
output:
  pdf_document: default
  html_notebook: default
---

```{r}
data <- read.csv(
  file = "returns.csv",
  sep = ";"
)
```

```{r}
data
```

## Returns and Log-Returns

In financial analysis, **returns** and **log-returns** play a key role in measuring the relative changes in asset prices over time.

### Simple Returns

A **simple return** calculates the relative change in price between consecutive periods:

$$
R_t = \frac{P_t - P_{t-1}}{P_{t-1}}
$$

where $P_t$ is the price at time $t$, and $P_{t-1}$ is the price in the previous period. Simple returns are intuitive, but they **do not add up linearly over time**, which can complicate multi-period analysis.

### Log-Returns

A **log-return** is defined as:

$$
r_t = \ln\frac{P_t}{P_{t-1}} = \ln(P_t) - \ln(P_{t-1})
$$

The advantage of log-returns is their **additivity over time** – the log-return over multiple periods is the sum of the individual period log-returns. They are also often more suitable for modeling financial data, especially when prices grow exponentially.

### Context in the Analyzed Data

In the data from **09/2023–02/2024**, there was a clear upward trend in the stock prices of both companies, which often moved **in parallel**. Therefore, a natural step is to analyze **log-returns**, which better highlight relative price changes and facilitate further statistical modeling.

Returns and log-returns were calculated in **Excel** and then imported into **R**, where **histograms** and a **scatter plot** comparing the log-returns of both companies were prepared. These visualizations allow for assessing the data distribution and the degree of interdependence between the companies.



```{r}
alior_log_returns <- data$daily_log_returns_alior[-1]
ing_log_returns <- data$daily_log_returns_ing[-1]
```

```{r}
# install.packages("ggplot2")
# install.packages("ggExtra")
library(ggplot2)
library(ggExtra)
```

```{r}
alior_ing <- data.frame(
  alior_log_returns = alior_log_returns,
  ing_log_returns   = ing_log_returns
)

p <- ggplot(alior_ing, aes(x = alior_log_returns, y = ing_log_returns)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  labs(
    title = "Alior & ING Bank Log-Returns (2023-2024) with Marginals",
    x = "Alior Bank",
    y = "ING Bank"
  ) +
  theme_minimal()

ggMarginal(
  p,
  type = "histogram",
  margins = "both",
  fill = "lightblue"
)
```


## Relationships Between Log-Returns

### Covariance

Covariance measures how two variables change together:

$$
\text{Cov}(X, Y) = \frac{1}{n-1} \sum_{i=1}^{n} (X_i - \bar{X})(Y_i - \bar{Y})
$$

- $X_i$, $Y_i$ – observations of variables X and Y  
- $\bar{X}$, $\bar{Y}$ – arithmetic means of the variables  
- $n$ – number of observations  

A positive covariance indicates that the log-returns of Alior and ING tend to move together – when one stock rises, the other usually rises as well. A negative covariance would suggest that an increase in one stock corresponds to a decrease in the other. A value close to zero indicates little or no linear relationship. In the banking sector, positive covariance is common, as stock prices often respond similarly to macroeconomic factors and monetary policy.

### Correlation

Correlation normalizes covariance, resulting in a value between $[-1,1]$, which makes it easier to assess the strength of the relationship.

#### Pearson Correlation (P)

$$
r_P = \frac{\text{Cov}(X, Y)}{\sigma_X \sigma_Y}
$$

Pearson correlation measures the linear relationship between Alior and ING log-returns. A value close to 1 means that changes in one stock’s log-returns are strongly associated with changes in the other in the same direction. A value close to -1 indicates a strong inverse relationship, and a value around 0 indicates little or no linear correlation. In our case, $r_P \approx 0.46$ shows a moderate positive linear relationship, meaning the stocks often move in the same direction but not perfectly in sync.

#### Spearman Correlation (S)

$$
r_S = 1 - \frac{6 \sum d_i^2}{n(n^2-1)}
$$

Spearman correlation is based on the ranks of observations and measures monotonic relationships, without assuming linearity. A positive value indicates that higher log-returns of one stock generally correspond to higher log-returns of the other, even if the relationship is not strictly linear. In practice, $r_S \approx 0.43$ confirms a moderate common trend in Alior and ING log-returns, consistent with expectations for the banking sector.



```{r}
cov(alior_log_returns, ing_log_returns)
cor(alior_log_returns, ing_log_returns, method = "pearson")
cor(alior_log_returns, ing_log_returns, method = "spearman")
```

### Analysis of Log-Return Relationships Between Alior and ING

Analyzing covariance and correlation allows us to assess the extent to which the log-returns of two companies move together.

- **Positive covariance** indicates that Alior and ING log-returns tend to rise together. In other words, when one stock increases, the other generally also shows an upward movement.  

- **Pearson correlation ($r \approx 0.46$)** shows a moderate positive linear relationship. This means that the stocks’ log-returns generally tend to move in the same direction, although some scatter exists and the relationship is not perfectly linear.  

- **Spearman correlation ($r \approx 0.43$)** measures monotonic association and indicates that higher ING log-returns are usually associated with higher Alior log-returns, even if the relationship is not exactly linear.  

In summary, both linear and monotonic analyses confirm a moderate positive dependence between Alior and ING log-returns, which is consistent with expectations for companies in the same banking sector.


## Linear Regression of Alior Log-Returns on ING

The purpose of linear regression is to create a model:

$$
Y = \beta_0 + \beta_1 X + \varepsilon
$$

- $\beta_0$ – intercept  
- $\beta_1$ – slope (coefficient)  
- $\varepsilon$ – random error  

The model allows us to **describe the trend of the dependent variable $Y$ as a function of the independent variable $X$** and to predict future values.

### $R^2$ (Coefficient of Determination)

$$
R^2 = 1 - \frac{\sum (Y_i - \hat{Y}_i)^2}{\sum (Y_i - \bar{Y})^2}
$$

- $\hat{Y}_i$ – predicted values from the model  
- $\bar{Y}$ – mean of observed values  

**Interpretation:**
- $R^2 \in [0,1]$  
- Explains the proportion of variance in the dependent variable that is accounted for by the independent variable in the model  
- Example: $R^2 = 0.8$ means 80% of the variance in $Y$ is explained by the linear model


```{r}
model <- lm(alior_log_returns ~ ing_log_returns)
summary(model)
```

```{r}
library(ggplot2)
library(ggExtra)
```

```{r}
p <- ggplot(alior_ing, aes(x = alior_log_returns, y = ing_log_returns)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Alior & ING Bank Log-Returns (2023-2024) with Marginals",
    x = "Alior Bank",
    y = "ING Bank"
  ) +
  theme_minimal()

ggMarginal(
  p,
  type = "histogram",
  margins = "both",
  fill = "lightblue"
)
```

Linear regression analysis allows us to assess to what extent the variability of Alior log-returns can be explained by ING log-returns.

- **Intercept (0.0009562)** – the average Alior log-return when ING log-return = 0. In practice, this means that if ING does not move, Alior also remains practically unchanged.  

- **Slope / coefficient for ING (0.5212)** – indicates that, on average, each 1-unit increase in ING log-return is associated with a 0.5212-unit increase in Alior log-return.  

- **$R^2 = 0.2151$** – the model explains about 21.5% of the variability in Alior log-returns based on ING. This represents a moderate level of fit, typical in finance where stock movements are largely random.  

- $R^2 = 0.215$ also suggests that over 78% of the variance in Alior log-returns is due to other factors, such as random market fluctuations, actions of other companies, or macroeconomic factors.  

In summary, the linear regression confirms a statistical relationship between the log-returns of the two companies, but it is not a strong relationship. The $R^2 = 0.215$ result is consistent with expectations for financial data.

